Using Bayesian networks to predict changes

### 2010-12-07 cs886-sarah/Paper.tex

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

@@

Changes

@@

\subsection{CMDBs and Change Sets}
A Configuration Management Database (CMDB) is useful in Enterprise IT Management (EITM) since it provides information about the various critical components in a
system including hardware, software, and services provided by the company. It records the configuration of these items, their change history, their incident
history, as well as the relationships between them. Each item stored in the CMDB is referred to as a Configuration Item (CI). Figure~\ref{fig:cmdbExample} shows
an example CMDB to illustrate the concepts of CIs and relationships. In this example, there are two software applications being hosted by two different web
servers. Both applications communicate with the same databases through a load balancer. Such a visualization allows an IT analyst to better understand the
system at hand.

A CMDB provides a basis for decision making processes such as Incident Management, Change Management, etc.  In this paper, we focus on the process of Change
Management, and in particular, on the problem of change set detection .A \textit{change} is the addition, modification, or removal of anything that could affect
on IT services. A poorly planned change may lead to a fault in the system. Accordingly, when one wants to change one CI is the system, other CIs that might
need to be changed as well must be correctly identified. The set of CIs that will need to be modified for the change to be complete is called a change set.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{cmdbExample.PNG}
\caption{An IT System Stored in a CMDB}
\label{fig:cmdbExample}
\end{figure}

@@

\subsubsection{Tools}

\subsubsection*{Weka}

\subsubsection*{Banjo}
Banjo is a tool written in Java which infers the structure of a Bayesian network given training data. It has two different search algorithms: Greedy and
Simulated Annealing. For either of these search algorithms, it has two methods of proposing a new edge. The first is the ``proposeRandomLocalMove''
which basically proposes a random addition or removal of an edge. The other is ``proposeAllLocalMoves'' which proposes all possible moves, and only keeps the
best one (?). Banjo uses the BDe metric to compute a network's score.

\subsubsection*{JavaBayes}

@@

\section{Models Used}
\label{sec:modelsused}

We experimented with several models to predict change sets. First, we just used Weka's BayesNet classifier to build the structure and estimate the CPTs of
the network. In the second model, we used Banjo to learn the structure of the network, and then we used Weka to learn the CPTs. In the third model, we imposed
the structure of the network from the relationships already existing between CIs in the CMDB. Finally, in the fourth model, we tried to assume that CIs not in
a change order are unobserved data, and therefore have missing values.

In order to be able to test things properly, we chose a small dataset to be able to build our model. We used observations from three months data from January 1,
2008 to March 31, 2008 to build the model. However, in such a short period, there were already 2,841 distinct CIs appearing and 2,229 observations (i.e. change
orders). Therefore, in each model, some data preprocessing was done before building the model since having a network with over 2000 variables was very hard to
deal with.

\subsection{Model 1}
\label{sec:model1}

\subsection{Model 2}
\label{sec:model2}

\subsubsection*{Data Preprocessing}

For this model, we first got a list of all the CIs that appeared more than 12 times (i.e were associated with 12 different change orders in our training
observations). This yielded 120 distinct CIs. Then, in order to slightly increase our variable space to include other related CIs, we found all the parent CIs
related to these 120 CIs from the CMDB perspective. However, we ignored three common, and not extremely meaningful relations, which are  ``supports'', ``is
location for'', and ``backs up''. Adding the related parent CIs, we now had a set of 241 CIs. We then went back, and iterated through the observations, and
kept only observations for CIs that either had more than 12 different change orders or were related to these CIs (i.e that occurred in the set of 241 CIs).
This, finally, gave us a set of 170 distinct CIs which we used for our model. Additionally, filtering out CIs meant filtering out some of the observations that
did not have any CIs satisfying our criteria. This led to us having 1305 observations instead of 2841 which was a more manegeable set.

\subsubsection*{Building the Model}

Given that we had our training set, the next step was to build the network which has two parts: building the structure, and estimating the CPTs. Unfortunately,
Weka mainly focused on classification and clustering which is not exactly the same problem we are tackling here. However, Weka did provide a functionality to
learn the CPT of a network given a certain structure. We, therefore, used Banjo to build the structure of the network (Banjo does not estimate CPTs), and then
used Weka to estimate the CPTs on the produced network. In both cases, the same observations file was used to ensure that its the same data. However, each tool
accepted the data in different formats, so some translation was necessary there. After Weka has added the CPT information to the graph, we used this as the
final Bayesian network on which to do our analysis.

\subsubsection*{Inference}

At this point, we have the Bayesian network ready, and we would like to perform Inference. More formally, given that a CI will change (our evidence), we want
to infer the probability that the other CIs in the network might change as well. Unfortunately, neither of the two tools previously used provide an Bayesian
inference engine. We, therefore, use JavaBayes in this step since it accepts the same ARFF format used by Weka. We using a one month testing set where we try
to predict all the change orders in April 2008 (883 reports). There was a total of XX change orders in that month. For each change order, we would take the
first CI as the initial CI to change, then we would set that as evidence, and update the beliefs of all the nodes (CIs) in the network using JavaBayes. We would
then loop on all the updated CIs, and add those that match our threshold criteria to the predicted change set. To simulate a real life scenario, we then checked
which of these predicted CIs actually lies in the target change set we are trying to predict. This is similar to an analyst accepting CIs into their change set.
These common CIs would then also be marked as observations so that we can predict now that I'm going to change B and C too, what else do I need to change.
Again, we would calculate the posterior probability, and continue doing so until there are no more common CIs. We would then calculate the recall, precision,
and F-measure accordingly.

\subsubsection*{Results}

\subsection{Model 3}

\subsubsection*{Data Preprocessing}

For this model, we only used the list of CIs appearing more than 12 times (120 distinct CIs).

\subsubsection*{Building the Model}

Using CIs in this list only, we constructed the structure of the Bayesian network by adding in CMDB relationships between the CIs. We stuck to this set only to
ensure that all the CIs in the network, are also present in the observations. Fourty-eight edges were added to the structure of the network. Using these 120
CIs, we had XXX observations. Again, we loaded the produced structure in Weka, and learned the CPTs using the observations from the three months. We then took
the produced Bayesian network, and used JavaBayes to perform the inference.

\subsubsection*{Inference}

\section{Building the model}
\label{sec:model}

\section{Experimental Work}
\label{sec:exp}

>>>>

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{url}

@@

Change is inevitable in any Information Technology(IT) system. New features are added, different configurations are used, upgrades are introduced and new
software and hardware are added. In a large system, such changes can cause other parts of the system to malfunction without the ability of the analyst to
foresee this side effect. Accordingly, we need to ensure that we consider the impact of all changes before they are implemented, and make sure we adjust the
potentially affected components accordingly while implementing the change. This leads to the process of change set detection which is finding all the
components of the system that need to be included in your change set where the change set is the set of components that need to be changed to avoid any side
effects.

Many of the large organizations with large and sophisticated IT systems employ a Configuration Management Database (CMDB) to help them manage these systems.
The CMDB keeps track of all the components in a system, how they are related, as well as all the changes that have occurred to them. Any component in the CMDB
is called a Configuration Item (CI). In previous work~\cite{nadi2010}, we mined the CMDB repository for historical co-changes, and used the mined correlations
to predict change sets. That is, given an initial CI that is going to change, we predict what other CIs might need to be changed as well based on its change
history with other CIs. We obtained really good results in terms of recall and precision (69.8\% and 88.5\% respectively).

In this paper, we wish to examine the same problem, but from a Bayesian perspective. Instead of looking at pairwise historical co-changes of CI, we would like
to consider all the changed CIs in the past to deduce the causality relations between them. That is, what is the probability that if A changes, B changes as
well. In that case, A would be a parent to B in a Bayesian network. Accordingly, we explore the different ways a Bayesian network can be constructed from the
data we have, and then test the predictions produced by querying this network to examine the obtained recall and precision. We hope that we can obtain better
results using Bayesian networks since they are not limited to pairwise comparisons.

The rest of this paper is organized as follows. BLA BLA BLA

@@

\subsection{CMDBs and Change Sets}
A Configuration Management Database (CMDB)~\cite{itil} is useful in Enterprise IT Management (EITM) since it provides information about the various critical
components in a
system including hardware, software, and services provided by the company. It records the configuration of these items, their change history, their incident
history, as well as the relationships between them. Each item stored in the CMDB is referred to as a Configuration Item (CI). Figure~\ref{fig:cmdbExample} shows
an example CMDB to illustrate the concepts of CIs and relationships.

A CMDB provides a basis for decision making processes such as Incident Management, Change Management, etc.  In this paper, we focus on the process of Change
Management, and in particular, on the problem of change set detection. A \textit{change} is the addition, modification, or removal of anything that could affect
on IT services. A poorly planned change may lead to a fault in the system. Accordingly, when one wants to change one CI is the system, other CIs that might
need to be changed as well must be correctly identified. The set of CIs that will need to be modified for the change to be complete is called a change set.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{cmdbExample.PNG}
\caption{An IT System Stored in a CMDB. Each item shown is a CI, and CIs are tied through different relationship types}
\label{fig:cmdbExample}
\end{figure}

@@

\subsection{Bayesian Network Tools}

\subsubsection{Weka}

WEKA~\cite{hall2009weka} is a toolbox for machine learning. It provides different algorithms for classification and clustering as well as other machine
learning problems. It also provides Bayesian learning techniques for classification problems. Additionally, it provides explorations tools for Bayesian
Networks, and allows the user to learn the structure and CPTs of a Bayesian Network. We will mainly be using WEKA for learning the CPTs in a fixed
structure network using the SimpleEstimator algorithm~\cite{written2005data} implemented there. This simple estimates the probability values on edges based on
the frequency values of each variable given its parents in the training set.

\subsubsection{Banjo}
Banjo~\cite{banjotool} is a tool written in Java which infers the structure of a Bayesian network given training data. It has two different search algorithms:
Greedy and Simulated Annealing. For either of these search algorithms, it has two methods of proposing a new edge. The first is the ``proposeRandomLocalMove''
which basically proposes a random addition or removal of an edge. The other is ``proposeAllLocalMoves'' which proposes all possible moves, and only keeps the
best one (?). Banjo uses the BDe metric to compute a network's score.

\subsubsection{JavaBayes}

\section{Constructing the Bayesian Network}
\label{sec:modelsused}

Before constructing the Bayesian Network, we first had to process and prepare the data we have. Then, in order to construct a Bayesian network to use for
predictions there are two steps involved. First, determining the structure of the actual network, and then estimating the Conditional Probability Tables
(CPT). Section~\ref{sec:data} first explains the data set available to us, and the data preprocessing involved before building the network.
Section~\ref{sec:structure} then explains the different techniques we experimented with to build the network structure. Section~\ref{sec:cpt} shows the last
step to build the network which is estimating the CPTs. Figure~\ref{fig:process} shows this overall process.

\subsection{Data Preprocessing}
\label{sec:data}

\subsubsection*{Data Size}
The original data set is from three years, and has 7,999 distinct CIs, and 27,305 change orders. This amount of data was infeasible to work with as no tool
could handle such a large amount of variables in a Bayesian network. For the purposes of this project, which is mainly to experiment with Bayesian techniques
for change set prediction,
it is sufficient to choose a small representative subset of the data. Accordingly, in order to be able to test things properly, we used observations from three
months data from January 1, 2008 to March 31, 2008 to build the model. However, even in such a short period, there were already 2,841 distinct CIs appearing and
2,229 observations (i.e. change orders). Therefore, we needed to perform further data preprocessing.

\begin{figure}[!t]
\centering
\includegraphics[width=7cm]{graphics/constructingmodel.png}
\caption{Building the Bayesian Network}
\label{fig:process}
\end{figure}

First, we removed all CIs that have changed less than 12 times within this time frame  (i.e were associated with 12 different change orders in our training
observations). There is no particular reason for choosing 12 as a cut off. It simply gave a feasible data set to deal with. This yielded 120 distinct CIs. Then,
in order to slightly increase our variable space to include other related CIs, we found all the parent CIs related to these 120 CIs from the CMDB perspective.
However, we ignored three common, and not extremely meaningful relations, which are  ``supports'', ``is location for'', and ``backs up''. Adding the related
parent CIs, we now had a set of 241 CIs. However, some of the added parent CIs may not be in the original set of 2,841 CIs. Accordingly, we just kept CIs tht
appeared more than 12 times or were in the set of related CIs. This provided us our final data set of 170 CIs which we use throughout our models for fair
comparison. Additionally, filtering out CIs meant filtering out some of the observations that did not have any CIs satisfying our criteria. This led to us
having 1,305 observations instead of 2,229 which was a more manegeable set. Accordingly, our training data set consisted of 170 variables (CIs), and 1,305
observations (change sets).

FIGURE FOR DATA PROCESSING

\subsubsection*{Data Format}
\label{sec:format}

In the CMDB, a Change Order has several fields including the requester, the assignee, the start date of the change, the description and the change set field
(called the `Configuration items' fields). Of the fields in a change order, we use only the change set field. The advantage of only using this one field is that
change sets are easy to extract and easy to understand. In terms of an observation, CIs were either marked as ``true'' or ``false'' indicating whether they
have changed or not. Each observation initially consisted of all the in our data set marked as ``false''. For each change order in the training set, we would
mark each CI appearing in the ``Configuration Items'' field as true, while all those not appearing were left as ``false''. Therefore, the set of training
data consisted of an observation entry for each change order with the CIs marked as ``true'' or ``false'' accordingly.

\subsection{Network Structure}
\label{sec:structure}
There were different ways in which we could estimate the structure of the network, and so we built a Bayesian Network using each technique.

\subsubsection{Model 1 (Banjo)}
\label{sec:model1}

For the first model, we used Banjo to learn the structure of the Bayesian network from the training data. We use the Greedy searching algorithm implemented
there. Banjo produced five different top-scoring networks (having the same score), and we simply chose one of them as the network to be used.

\subsubsection{Model 2 (CMDB Relationships)}

For the second model, we used the relations existing in the CMDB to infer the structure of the network (that is place the edges between the nodes). We used two
variations for this. The first was to follow the direction of the relationship edges in the CMDB. That is if there is an edge from A to B, we will place an
edge in the Bayesian network from A to B (we will call this model 2A). The second was to reverse the direction of the relationship edges in the CMDB. That is,
if there is an edge from A to B, we will place an edge in the Bayesian network from B to A (we will call this model 2B). In both cases, however, we chose to
ignore some relationship edges (WHY, TRY WITHOUT). There was one problem, however, with the generated network. Two of the CIs had more than 15 parents to them
which means that their CPTs will be intractable to compute. For those two CIs, we simply removed all their parents to make the computation tractable.

\subsubsection{Model 3 (Missing Data)}

This model is the same as the second model, except that we experimented with the missing data feature in Weka. Weka allows the user to specify missing values
for any variable by simply putting a '?' instead of its value. Therefore, instead of putting the CIs that did not appear in a change order as false, we put a
'?' in their place. This seemed closer to practice, because in reality, we are not sure whether this CI actually changed or not. Therefore, this model also has
two parts: model 3A and model 3B where the first uses the CMDB relationships in their same direction, and the second uses the reverse direction.

\subsection{Estimating the CPTs}
\label{sec:cpt}

However, for all
models, we used the SimpleEstimator algorithm~\cite{written2005data} built in Weka to calculate the
CPTs. After setting the data set to be the observations in our training set, Weka
learned the CPTs for the network produced by Banjo. This complete network was then saved as an BIF XML file to be used by JavaBayes for general inference.

\section{Experiment Setup}

\begin{algorithm}[tb]
   \caption{Generating Predictions using the Bayesian Network}
   \label{alg:simulation}
\begin{algorithmic}
   \STATE {\bfseries Input:} changeOrder
    \STATE {\bfseries Input:} BayesianNetwork
    \STATE {\bfseries Input:} threshold
    \STATE Set initialCI = first CI in changeOrder
    \STATE Set occurredSet = changeOrder - initialCI
    \STATE Initialize observedSet = {firstCI}
    \STATE Initialize predictedSet = {}
   \REPEAT  
    \STATE Initialize newPredictions = {}
    \FOR{$node$ {\bfseries to} $BayesianNetwork$}
    \STATE posterior = perform inference using Banjo
    \IF{$posterior > threshold$} 
      \STATE Add node to newPredictions   
    \ENDIF
    \ENDFOR

  \FOR{$prediction$ {\bfseries to} $newPredictions$}
    \IF{$prediction  occurredSet$}
      \STATE Add prediction to observedSet
    \ENDIF
    \ENDFOR
\STATE Add newPredictions to predictedSet

   \UNTIL{$newPredictions$ is $empty$}
\end{algorithmic}
\end{algorithm}

\subsection{Change Set Detection Process Simulation}

The change set detection process provided is an iterative, collaborative process between the tool and the analyst. When the tool suggests CIs, the analyst can
accept or reject these CIs (i.e add them to the change set or not). Based on the CIs added to the change set, the analyst can ask the tool for more
suggestions. This is along the lines of ``Now, that I am also going to change these CIs, what else do I need to change?''.

To simulate this process for our experiments, we do the following. At this point, we have the Bayesian network ready, and we would like to perform Inference.
More formally, given that a CI will change (our observation), we want to infer the probability that the other CIs in the network might change as well.
Unfortunately, neither of the two tools previously used provide an Bayesian inference engine. We, therefore, use JavaBayes in this step since it accepts the
same ARFF format used by Weka. We using a one month testing set where we try to predict all the change orders in April 2008. There was a total of
883 change orders in that month.

For each change order, we would take the first CI as the initial CI to change, then we would set that as an observed node, and update
the beliefs of all the nodes (CIs) in the network using JavaBayes. We would then loop on all the updated CIs, and add those that match our threshold criteria to
the predicted change set. To simulate a real life scenario, we then checked which of these predicted CIs actually lies in the target change set we are trying to
predict. This is similar to an analyst accepting CIs into their change set. These common CIs would then also be marked as observations so that we can predict
now that I'm going to change these accepted CIs as well, what else do I need to change. Again, we would calculate the posterior probability, and continue doing
so until there are no more common CIs. All the CIs that match the threshold criteria (whether accepted by the analyst or not) are part of the predicted
set.

@@

Algorithm~\ref{alg:simulation} shows this process, and how the recall and precision are calculated accordingly.

\section{Results}
\label{sec:results}
